Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-18 WholePlotTrajectories.tex
Results & appendix

The large areas covered with tropical forests worldwide hold crucial
economic, social and cultural value. They provide wood and multiple
non-timber forest products, shelter a diversified fauna, regulate the
local climate, support the carbon, water and nutrient cycles, and ensure
cultural and human well-being. The simultaneous increase of forests
products demand and substantial climatic changes currently heighten the
pressure on the remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}. It threatens the maintenance
of communities structure, composition and functioning and the underlying
dynamics in space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are contantly re-shaped by
the natural disturbance events changing both abiotic environment,
through the fluxes of light, heat and water, and biotic interactions and
competitive pressure \citep{Goulamoussene2017}. The cornerstone of
tropical forests ecology is to understand the underlyings of ecosystem
response to disturbance, its mechanisms and drivers
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. Consistent models assessed these
structural parameters and highlighted the maintenance of ecosystems
processes and services \citep{Denslow2000, Blanc2009}. The response of
tree diversity, however, remains unclear albeit it determines ecosystems
productivity, stability and functioning
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by post-disturbance environmental changes
\citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Validations of the IDH in
the long term, however, remain scarce and mainly rely on the analysis of
the rough species richness that gives limited or misleading information
on forests recovery and functioning \citep{Martin2015, Chaudhary2016}.
More releveant monitoring would encompass communities composition, that
is crucial for conservation issues, and diversity profiles, that reveals
ecological rules underlying communities' structure
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, account
for species biological attributes and role the ecosystem through a
functional approach assesses communities functioning and corresponding
mechanisms
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect major functional traits related to species ecology and
performance were largely adopted \citep{Diaz2005, Villeger2008a}
highlighting, for example, the environmental filters fostering
disturbance resistant species with rapid growth and efficient resources
acquisition in tropical rainforests \citep{Molino2001, Haddad2008}. In
disturbed tropical rainforests, this framework translates into the
fostering of disturbance resistant species with rapid growth and
efficient resources acquisition \citep{Molino2001, Haddad2008}. It
results in a shift from ``conservative'' slow-growing species dealing
with scarce resources that dominate before disturbance, to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. It was
mirrored by shifts in key functional traits related to resource
acquisition (leaf area, density and chlorophyll content, and stem
specific gravity and bark thickness), tree growth and reproduction life
history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

Functional and taxonomic trajectories relate to distinct characteristics
and community processes, and might therefore be decoupled as already
observed in tropical forests \citep{Lohbeck2015, Guariguata2001}. The
combination of taxonomic and functional characteristics is translated by
the functional redundancy, measuring the amount of species sharing same
trait values \citep{Carmona2016}. Functional redundancy is a major
determinant of communities resilience as high redundancy, like in highly
diverse tropical forests \citep{Bellwood2006}, mitigates the impacts of
species removal on ecosystem functioning
\citep{Trenbath1999, Elmqvist2003, Diaz2005}. It is therefore a key
element as a link between taxonomic and functional diversities and as a
determinant of communities resilience.

Grasping all facets of communities response to disturbance then requires
to disentangle the taxonomic and functional trajectories in diversity
and composition. These trajectories would highlight communities
resilience and the drivers of their response to disturbance, clarifying
the tenants of the IDH in the long term for tropical forests, that are
as much insights for future adaptive conservation strategies
\citep{Adler2007}. Here we monitored over 30 years the response of 75 ha
of forests plots set up on a gradient of disturbance intensity, from 10
to 60\% of ecosystem biomass removed. We made use of a large functional
traits database browsing major leaf, stem and seed traits and species
maximum height to draw the trajectories over time of communities
taxonomic and functional composition and diversity. Specifically, we (i)
highlighted the determinant of communities functoinal and taxonomic
response to disturbance, (ii) tested the validity of the IDH in the long
term for tropical forest and (iii) questioned the completeness of
communities recovery regarding their functional redundancy.

@@

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a
lowland tropical rain forest corresponding to a tropical wet climate
with mean annual precipitation averaging 2980 mm.y\textsuperscript{-1}
(30-y period) with a 3-month dry season (\textless{} 100
mm.month\textsuperscript{-1}) from mid-August to mid-November, and a
one-month dry season in March \citep{Wagner2011}. Elevation ranges
between 5 and 50 m and mean annual temperature is 26°C. Soils correspond
to thin acrisols over a layer of transformed saprolite with low
permeability generating lateral drainage during heavy rains. The
experiment corresponds to a network of twelve 6.25ha plots that have
undergone a gradient of three logging, thinning and fuelwood treatments
(Table \ref{tab:Tab1}). Disturbance treatments were attributed according
to a randomized plot design with three replicate blocks of four plots.
The disturbance corresponds to averages of 10 trees removed per hectare
with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1),
32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40
cm DBH for treatment 3 (T3). Treatments T2 and T3 besides included the
thinning of trees by poison girdling \citep{Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance (ref to be found). Biomass measurements were performed with
the BIOMASS package R package \citep{Biomass2018}.

@@

\subsection{Inventories protocol and dataset
collection}\label{inventories-protocol-and-dataset-collection}

The study site corresponds to a tropical rainforest with a dominance of
Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical
families. In the twelve experimental plots of the experiment, all trees
above 10 cm DBH were mapped and measured annually since 1984. During
inventories, trees were first identified with a vernacular name assigned
by the field team, and afterward with a scientific name assigned by a
botanist during regular botanical campaigns. In 1984, specific
vernacular names were given to 62 commercial or common species whereas
other less common species were identified under two identifiers only
separating trees and palm trees. The botanical campaigns carried every 5
to 6 years to identify all trees at the species level only started in
2003 and identification practices varied among plots and successive
campaigns. This raised methodological issues as vernacular names usually
correspond to different botanical species, resulting in significant
taxonomic uncertainties that had to be propagated to composition and
diversity metrics through a Bayesian framework. The uncertainty
propagation was done by the replenishment of inventories completed at
genus level from real incomplete ones on the basis of
vernacular/botanical names association.

Vernacular names were replaced through multinomial trials
\(M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)\)
based on the observed association probability
\(\big[\alpha_1, \alpha_2,…, \alpha_3\big]\) between each vernacular
name \emph{v} and the species \(\big[s_1, s_2, …, s_N\big]\) recorded in
the inventory. See appendix 1 and \citet{Aubry-Kientz2013} for the
detailed methodology.

The functional diversity metrics used a dataset for 6 functional traits
representing leaf economics (leaves thickness, toughness, total
chlorophyll content and specific leaf area, the leaf area per unit dry
mass) and wood economics (wood specific gravity and bark thickness), and
life history traits (maximum specific height and seed mass).

The trait database came from the BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}
where trait values were assessed from a selection of individuals located
in nine permanent plots in French Guiana, including two in Paracou.
Missing trait values were filled using multivariate imputation by
chained equation (mice) restricted to samples pertaining to the next
higher taxonomic level, in order to account for the phylogenetic signal
of the functional traits. The dataset comprised 294 botanical species
pertaining to 157 botanical genera.

Whenever a species inventoried was not in the dataset, it was attributed
a set of traits values randomly sampled among species of the same next
higher taxonomic level. As seed mass information corresponds to a
classification into mass classes, no data filling process was applied so
analysis were performed considering the 414 botanical species of the
seed mass dataset. All composition and diversity metrics corresponded to
the average obtained after 50 iterations of the taxonomic uncertainty
propagation framework and of the filling process of missing trait
values.

@@

To counter the remaining taxonomic uncertainty, plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after
disturbance was visualized by their trajectories over 30 years in a
two-dimensional ordination space. Two NMDS were conducted, either from
taxonomic flora inventories or from plots functional composition based
on the 7 leaf, stem and life history traits (without seed mass classes).
In both cases the NMDS were performed using occurrence-based (Jaccard)
as well as abundance-based (Bray-Curtis) dissimilarity measures.
Trajectories along time were reported through the euclidean distance of
successive inventories to the reference inventories in 1989, 5 years
after disturbance, when the uncertainty degree did not exceed 30\% of
undetermined trees. The trajectories of the leaf and stem and life
traits were also visualized with the community weighted means (CWM),
representing the average trait value in a community weighted by relative
abundance of the species carrying each value
\citep{Diaz2007, Garnier2004}. To compensate the intrinsinc difference
among plots the trajectories corresponded to the differences along time
with the reference inventory in 1989. Species seed mass corresponded to
5 classes of increasing mass, seed mass trajectories were therefore
reported as the proportion of each class in the inventories.

@@

The functional redundancy is measured as the overlap among species in
communities' functional space \citep{Carmona2016}. First, the samples of
the trait database are mapped in a 2-dimensional plan with a PCA
analysis. Then, multivariate kernel density estimator associated with
individual trees were give species traits probability distribution
(TDP). TDP, weighted by species abundance, are summed to give
communities TDP: the functional redundancy is the sum of TDP overlaps,
corresponding to the average number of species that could be removed
from without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

Communities taxonomic diversity trajectories were examined through the
Richness, Shannon and Simpson diversities at genus level, in relation to
the 1989 inventories (5 years after disturbance) (See annexe I). For
undisturbed plots the Richness, Shannon and Simpson diversities remained
stable over the 30 years monitored. After low disturbance intensity the
richness increased, reaching a maximum gain of 14 botanical genera (plot
3 from treatment 2). After intense disturbance, plots taxonomic richness
followed unimodal trajectories first decrease for around ten years and
then increasing to return to pre-disturbance values. For all disturbed
plots the taxonomic evenness (Shannon and Simpson diversities)
increased, following unimodal trajectories with a just beginning return
towards initial values after 30 years. The maximum evenness, reached
after around 20 years, was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Only two T3 plots (plots 8 and 12)
were still increasing 30 years after disturbance \ref{fig:IDHplot},
suggesting similar but delayed trajectories.

Communities functional diversity trajectories were examined through the
Rao diversity based on the 7 leaf, stem and life history traits (the
seed mass was excluded as a qualitative variable). The plot 7 from
treatment 1 displayed a constantly outlying diversity and was removed
from the graphical representation for better readability (see appendix
for full graphs). The functional diversity of all undisturbed
communities remained stable along the 30 years. For all disturbed
communities the functional diversity followed unimodal trajectories with
a return towards the initial state, comparable to the control plots.

The impact of disturbance was examined specifically through the linear
correlation between the disturbance intensity (intial \%AGB removed) and
the Simpson and Rao diversities (diversities of order 2) at three focus
times after disturbance (10, 20 and 30 years) \ref{fig:IDHplot}. The
Simpson diversity was weakly related to the disturbance intensity
(\(R^2<0.25\)) and only from 20 years after disturbance. The Rao
diversity was more strongly correlated to the disturbance intensity
(\(0.55<R^2<0.75\)) for the three times studied. The impact of
disturbance, translated by the slope of the linear correlation, was the
highest 20 years after disturbance.

@@

In the inventories from 1989 (5 years after disturbance) to 2015 (31
years after disturbance), 828388 trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded. The
trajectories of taxonomic and functional composition were visualized in
a two dimensional ordination space mapping the successive inventories
according to their flora and the corresponding traits. Classifications
were performed using either abundance-based Bray-Curtis (Figure
\ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity. Both
metrics gave similar results, so only analysis using Bray-Curtis
dissimilarity will be discussed here.

While both taxonomic and functional composition remained stable in
undistrubed communities, after disturbed they followed consistent
trajectories over time that revealed significant compositional changes.
Considering the mapping of functional traits in the same dimensional
space, the post-disturbance trajectories corresponded to shifts towards
more acquisitive functional strategies: disturbed communities changed
from high average WD to high average SLA and chlorophyll content. For
disturbed communities the distance of successive inventories to the
reference inventory in 1989 followed unimodal trajectories (AppendixI,
Figure A2). The maximum of communities dissimilarity to their initial
state was positively correlated to the disturbance intensity for both
taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.91\) and
\(\rho_{spearman}^{functional}=0.96\) respectively). The time at maximum
dissimilarity was reached around 26 years after disturbance for
taxonomic composition and 22 years for functional composition. All
trajectories followed cyclic composition changes with a return towards
the initial composition (Figure \ref{fig:NMDSplans}).

@@

The changes observed in plots functional composition went hand to hand
with consistent trajectories of the 8 functional and life history traits
visualized through plots community weighted means (CWM) (Figure
\ref{fig:CWM}) of the leaves thickness, chlorophyll content, toughness
and specific area, wood specific gravity, bark thickness, seed mass and
maximum adult height.

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots, all traits and seed mass proportions followed a
unimodal trajectories and either stabilized or returned towards their
initial values 30 years after disturbance. At that time the weighted
means of communities specific maximum height at adult stage
(\emph{Hmax}), leaf toughness (\emph{L\_toughness}) and wood specific
gravity (\emph{WD}) remained significantly lower than their initial
value and than these of the control plots (Figure \ref{fig:CWM}). The
weighted means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value,
similar as those of undisturbed plots. For all traits the maximum
difference to initial state was correlated to the disturbance intensity
(\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

Communities functional redundancy was measured as the sum of weighted
overlap among species in communities' functional space. Functional
spaces were defined in the main two-dimensional space of a PCA analysis
summarizing the 7 leaf, stem, and maximum height traits (see appendix I
for PCA details). Communities functional redundancy remained stable in
control plots but after disturbance the redundancy trajectories were
quite variable (See appendix I). Globally after most intense disturbance
(plots T2 and T3) communities redundancy decreased at first place before
increasing to edge, recover or exceed the initial value.

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped-shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss observed was positively correlated with the
disturbance intensity (XX spearman to be measured) and none of the
disturbed plots had recovered the functional redundancy within the
initial functional space.

>>>>

The large areas covered with tropical forests worldwide hold crucial
economic, social and cultural values. They provide wood and multiple
non-timber forest products, shelter a diversified fauna, regulate the
local climate, support the carbon, water and nutrient cycles, and ensure
cultural and human well-being. The simultaneous increase of forests
products demand and the substantial climatic changes currently heighten
the pressure on remaining forests
\citep{Gibson2011a, Morales-Hidalgo2015}, threatening the maintenance of
communities structure, composition and functioning and their dynamics in
space and time \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forest, ecological communities are constantly re-shaped by
the natural disturbance events changing both abiotic environment,
through the fluxes of light, heat and water, and biotic interactions and
competitive pressure \citep{Goulamoussene2017}. The cornerstone of
tropical forests ecology is to understand the mechanisms and the
determinants of ecosystems response to disturbance
\citep{White2001, Chazdon2003a}. For now, this has been largely studied
through structural parameters, rapid and convenient to measure, as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016}. These structural parameters were
thereafter consistently modeled and gave important insights into the
maintenance of ecosystems processes and services
\citep{Denslow2000, Blanc2009}. However the response of forests
diversity in tree species remains unclear, albeit it determines the
productivity, stability and functioning of ecosystems
\citep[\citet{Liang2016}]{Tilman2014} and would be most probably
impacted by the changes following disturbance \citep{Baraloto2012a}.

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
releveant monitoring would encompass communities composition, crucial
for conservation issues, and complete diversity profiles encompassing
species evenness in addition to their richness, that reveals ecological
rules shaping communities
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, account
for species biological attributes and role in the ecosystem is essential
to understand the correlations between ecosystems biodiversity,
functioning and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect functional approaches based on major traits related to species
ecology and performance was largely adopted
\citep{Diaz2005, Villeger2008a}, for example highlighting in tropical
rainforests the environmental filters fostering fast growing species
with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008}. Post-disturbance trajectories then
corresponded to a shift from ``conservative'' slow-growing species
dealing with scarce resources dominating before disturbance, to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. This
shift translated into consistent trajectories for key functional traits
related to resource acquisition (leaf area, density and chlorophyll
content, and stem specific gravity and bark thickness), tree growth and
reproduction and life history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

Functional and taxonomic trajectories are both essentiall to fully
assess communities response to disturbance. Both relate to distinct
ecological rules and processes and might then be decoupled as it was
observed already in tropical forests
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary, they are combine into communities
functional redundancy that measures the amount of species sharing same
trait values \citep{Carmona2016}. Functional redundancy then explicits
the link between otherwise decoupled taxonomic and functional
diversities and is central for communities description. Besides,
functional redundancy determines communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Grasp all facets of communities response to disturbance amount to
disentangle the taxonomic and functional trajectories in diversity and
composition. These trajectories would highlight the ecological rules
constraining, or not, communities dynamics towards a recovery of initial
composition, diversity and functioning. They would therefore clarify the
tenants in the long term of the debated Intermediate Disturbance
Hypothesis for tropical forests, which would be as much insights for
future adaptive conservation strategies \citep{Adler2007}. Here we
monitored over 30 years the response of 75 ha of forests plots set up on
a gradient of disturbance intensity, from 10 to 60\% of ecosystem
biomass removed. We made use of a large functional traits database
browsing major leaf, stem and seed traits and species maximum height to
draw the trajectories over time of communities taxonomic and functional
composition and diversity. Specifically, we (i) questioned the coupling
between taxonomic and functional response to disturbance and identified
the underlying assembly processes, which allowed to (ii) clarify the
validity of the IDH in the long term for tropical forest and (iii)
question the completeness of communities recovery regarding their
functional redundancy.

@@

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a
lowland tropical rain forest in a tropical wet climate with mean annual
temperature of 26°C, mean annual precipitation averaging 2980
mm.y\textsuperscript{-1} (30-y period) and a 3-month dry season
(\textless{} 100 mm.month\textsuperscript{-1}) from mid-August to
mid-November, and a one-month dry season in March \citep{Wagner2011}.
Elevation ranges between 5 and 50 m and soils correspond to thin
acrisols over a layer of transformed saprolite with low permeability
generating lateral drainage during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}). Disturbance treatments were attributed according
to a randomized plot design with three replicate blocks of four plots.
The disturbance corresponds to averages of 10 trees removed per hectare
with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1),
32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40
cm DBH for treatment 3 (T3). Treatments T2 and T3 besides included the
thinning of trees by poison girdling \citep{Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance (ref to be found) measured with the BIOMASS R package
\citep{Biomass2018}.

@@

\subsection{Inventories protocol and dataset
collection}\label{protocols}

The study site corresponds to a tropical rainforest with a dominance of
Fabaceae, Chrysobalanaceae, Lecythidaceae and Sapotaceae botanical
families. In the twelve experimental plots of the experiment, all trees
above 10 cm DBH are mapped and measured annually since 1984. Trees are
first identified during inventories with a vernacular name assigned by
the field team, and afterward with a scientific name assigned by a
botanist during regular botanical campaigns. In 1984, specific
vernacular names are given to 62 commercial or common species whereas
more infrequent ones were identified under general identifiers only
distinguiching trees and palm trees. From 2003 botanical campaigns were
conducted every 5 to 6 years to identify all trees at the species level
but identification practices still varied among plots and campaigns.

This variability of protocols raised methodological issues as vernacular
names usually correspond to different botanical species. It resulted in
significant taxonomic uncertainties that had to be propagated to
composition and diversity metrics. The uncertainty propagation was done
through a Bayesian framework reconstituting complete inventories at
genus level from real incomplete ones on the basis of
vernacular/botanical names association. Vernacular names were replaced
through multinomial trials
\(M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big)\)
based on the association probability
\(\big[\alpha_1, \alpha_2,…, \alpha_3\big]\) observed across all
inventories between each vernacular name \emph{v} and the species
\(\big[s_1, s_2, …, s_N\big]\). See appendix 1 and
\citet{Aubry-Kientz2013} for the detailed methodology.

The functional approach used a dataset of 6 functional traits
representing leaf economics (leaves thickness, toughness, total
chlorophyll content and specific leaf area, the leaf area per unit dry
mass) and wood economics (wood specific gravity and bark thickness), and
life history traits (maximum specific height and seed mass). The trait
database came from the BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}
where trait values were assessed from a selection of individuals located
in nine permanent plots in French Guiana, including two in Paracou, and
comprised 294 botanical species pertaining to 157 botanical genera.
Missing trait values were filled using multivariate imputation by
chained equation (mice) from the mice R package \citep{Mice2011}.
Imputations were restricted within genus, or family when samples wre too
scarce, in order to account for the phylogenetic signal of the
functional traits. Whenever a species inventoried was not in the
dataset, it was attributed a set of traits values randomly sampled among
species of the same next higher taxonomic level (same genus or family).
As seed mass information corresponds to a classification into mass
classes, no data filling process was applied and analysis were
restricted to the 414 botanical species recorded.

All composition and diversity metrics corresponded to the average
obtained after 50 iterations of the taxonomic uncertainty propagation
framework and of the filling process of missing trait values.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification protocols (see {[}\#protocols{]}), the taxonomic
composition and diversity analysis were conducted at the genus level,
\emph{i.e.} referring to the genus of observed or trialed botanical
names. Trajectories of communities taxonomic and functional variations
in composition after disturbance were followed in a two-dimensional
ordination space the 30 years monitored. Two NMDS were conducted to map
either taxonomic flora inventories or communities functional composition
based on the 7 leaf, stem and life history traits (without seed mass
classes). In both cases the NMDS were performed using occurrence-based
(Jaccard) and abundance-based (Bray-Curtis) dissimilarity measures.
Trajectories along time in the plan were reported through the euclidean
distance of successive inventories to the reference inventories in 1989,
5 years after disturbance, when the uncertainty degree did not exceed
30\% of undetermined trees. The trajectories of the leaf and stem and
life traits were also visualized with the community weighted means
(CWM), representing the average trait value in a community weighted by
relative abundance of the species carrying each value
\citep{Diaz2007, Garnier2004}. To compensate the intrinsinc difference
among plots the trajectories corresponded to the differences along time
with the reference inventory in 1989. Species seed mass corresponded to
5 classes of increasing mass, seed mass trajectories were therefore
reported as the proportion of each class in the inventories.

@@

The impacts of initial disturbance were first tested with the spearman
rank correlation between the extremum of taxonomic and functional
metrics reached over the 30 years and the initial \%AGB removed. Then
they were analysed through the linear correlations between Simpson and
Rao diversities and the initial \%AGB removed at 10, 20 and 30 years
after disturbance.

The functional redundancy was measured as the overlap among species in
communities' functional space \citep{Carmona2016}. The samples of the
trait database were first mapped in a 2-dimensional plan from a PCA
analysis. Then, multivariate kernel density estimator associated with
individual trees returned species traits probability distribution (TDP).
Species TDP weighted by species abundance were eventually summed for
each community: the functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

In the inventories from 1989 (5 years after disturbance) to 2015 (31
years after disturbance), 828388 trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded.
Communities taxonomic diversity trajectories were examined through the
Richness, Shannon and Simpson diversities at genus level, in relation to
the 1989 inventories (5 years after disturbance) (See annexe I). For
undisturbed plots the Richness, Shannon and Simpson diversities remained
stable over the 30 years monitored. In disturbed communities the
taxonomic richness increased after low disturbance intensity, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2) while it
followed unimodal trajectories after intense disturbance, decreasing for
ten years before recovering pre-disturbance values. In all disturbed
plots the taxonomic evenness (Shannon and Simpson diversities)
increased, following unimodal trajectories with a maximum, reached after
around 20 years, positively correlated to the disturbance treatment
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Return towards initial evenness
values was beginning after 30 years except for two T3 plots (plots 8 and
12) which evenness still increased, suggesting similar but delayed
trajectories \ref{fig:IDHplot}.

Trajectories of communities functional diversity were examined through
the Rao diversity based on the 7 leaf, stem and life history traits (to
the exception of seed mass). The plot 7 from treatment 1 displayed a
constantly outlying diversity and was removed from the graphical
representation for better readability (see appendix for full graphs). In
undisturbed plots the functional diversity remained stable along the 30
years while in disturbed plots it followed unimodal trajectories with a
return towards initial values that strated around 20 years after
disturbance.

The impact of disturbance was examined specifically through the linear
correlation between the intial \%AGB removed and the Simpson and Rao
diversities (diversities of order 2) after 10, 20 and 30 years
\ref{fig:IDHplot}. The correlation with disturbance intensity was weak
for the Simpson diversity (\(R^2<0.25\)) and only valid from 20 years
after disturbance but it was much stronger for the Rao diversity
(\(0.55<R^2<0.75\)) for all the time studied. Slope of linear
correlations, reflecting the impact of disturbance, was the highest 20
years after disturbance.

@@

The trajectories of taxonomic and functional composition were visualized
in a two dimensional ordination space mapping the successive inventories
according to their flora and corresponding traits. Classifications were
performed using either abundance-based Bray-Curtis (Figure
\ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity, both
giving similar results only analysis using Bray-Curtis dissimilarity are
discussed.

While both taxonomic and functional composition remained stable in
undistrubed communities, they followed consistent trajectories over time
after disturbance which revealed significant compositional changes.
According to the mapping of functional traits (see appendix I) these
compositional changes corresponded to shifts towards species with more
acquisitive functional strategies, from communities with high average WD
to high average SLA and chlorophyll content. For disturbed communities
the distance of successive inventories to the 1989 reference inventory
followed unimodal trajectories translating cyclic compositional changes
with a recovery of the initial composition (Figure \ref{fig:NMDSplans}).
The maximum dissimilarity with the initial state was positively
correlated to the disturbance treatment for both taxonomic and
functional composition (\(\rho_{spearman}^{taxonomic}=0.91\) and
\(\rho_{spearman}^{functional}=0.96\) respectively) and the time at
maximum was reached around 26 years after disturbance for taxonomic
composition and 22 years for functional composition.

@@

The changes observed in plots functional composition went hand to hand
with consistent trajectories of the 8 functional and life history traits
visualized with the trajectories of community weighted means (CWM) of
leaves economics (leaves thickness, chlorophyll content, toughness and
specific area), wood economics (wood specific gravity, bark thickness),
and life history traits (seed mass and maximum adult height) (Figure
\ref{fig:CWM}).

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions followed unimodal trajectories either stabilizing or
returning towards their initial values. Thirty years after disturbance
the weighted means of communities specific maximum height at adult stage
(\emph{Hmax}), leaf toughness (\emph{L\_toughness}) and wood specific
gravity (\emph{WD}) remained significantly lower than their initial
value (Figure \ref{fig:CWM}). The weighted means of bark thickness
(\emph{Bark\_thick}) similarly remained substantially higher than
initially for all disturbed plots while the specfic leaf area
(\emph{SLA}) had almost recovered its initial value. For all traits the
maximum difference to initial state was correlated to the disturbance
intensity (\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

Communities functional redundancy was measured as the sum within
communities the species weighted functional overlap based on the 7 leaf,
stem, and maximum height traits (see appendix I for PCA details).
Communities functional redundancy remained stable in control plots but
after disturbance the redundancy trajectories were quite variable (See
appendix I) and apparently independently of the initial disturbance.
Globally after most intense disturbance (plots T2 and T3) communities
redundancy decreased at first place before increasing to edge, recover
or exceed the initial value.

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped-shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss was positively correlated with the disturbance
intensity (XX spearman to be measured) and the initial value had not
recovered for any disturbed communities.
